A Large-scale Study of Polymorphism in R

### 2019-04-05 notebooks/data_builder.Rmd
last plot, modified discussion, still plugging away at it

@@

```{r}
# first, turn specials, closures, and builtins to functions
psp <- transform_new_df_w_TS_map(psp, type_map_to_fun)
```

@@

@@

data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L1_funs_c/1000, tot_L1_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L1_funs_c/tot_L1_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)
```

@@

```{r}
# L2 Time
psp_L2 <- read_csv("../data/partial_L1_L2.csv.gz")
psp_L2 <- psp_L2[!psp_L2$pkg == "SUBTYPE",]

# L2 collapsed?
# what <- read_csv("../data/partial_L1_sanitized_class_collapsed.csv.gz")
what <- read_csv("../data/partial_L1_collapsed_L2_new.csv.gz")
```

>>>>

WORKFLOW
0. build data on prl3
  0.1. reduce traces
  0.2. make_new_df \forall the packages
  0.3. rbindlist them all
1. get data from prl3 and do the notebook up to 2.

@@

```{r}
# first, turn specials, closures, and builtins to functions
psp <- transform_new_df_w_TS_map(psp, type_map_to_fun)

# any other transformations?
```

@@

WORKFLOW
2. write the partials out to csv for java
3. in java, call doIt (remember -L=L0 for L0, -L=L1 for L1) ALSO NOTE FILE NAMES

@@

data.frame(`Signature Type`=c("Monomorphic", "Total Seen"), Count=paste0(c(mono_L1_funs_c/1000, tot_L1_funs/1000) %>% as.integer, "K"), Percentage=c(paste0(round(mono_L1_funs_c/tot_L1_funs * 100, 1), "%"), "---")) %>% xtable %>% print(include.rownames=F)

# actually want plots, next
```

```{r}
# Actually, want per fun plots
pf_L0 %>% mutate(num_sigs = 1) %>% group_by(pkgfun) %>% summarize(num_sigs=sum(num_sigs)) -> plot_L0
pf_L1 %>% mutate(num_sigs = 1) %>% group_by(pkgfun) %>% summarize(num_sigs=sum(num_sigs)) -> plot_L1

plot_L0 %>% arrange(num_sigs) %>% mutate(pkgfun_o = factor(pkgfun, pkgfun)) -> plot_L0_df
plot_L1 %>% arrange(num_sigs) %>% mutate(pkgfun_o = factor(pkgfun, pkgfun)) -> plot_L1_df

# if L0 has no data entry, get rid of it
if (plot_L0_df[plot_L0_df$pkgfun == "no_data::no_data",] %>% nrow == 1)
  plot_L0_df <- plot_L0_df[!plot_L0_df$pkgfun == "no_data::no_data",]

L0_perc_poly <- round(100 * nrow(plot_L0_df[plot_L0_df$num_sigs > 1, ] )/nrow(plot_L0_df), 1)
L1_perc_poly <- round(100 * nrow(plot_L1_df[plot_L1_df$num_sigs > 1, ] )/nrow(plot_L1_df), 1)

# maybe only plot those with more than 1 sig?
plot_L0_df %>%
  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
  geom_line() + labs(title = "Number of Function Signatures per Function (L0)",
                                    x = "",
                                    y = "Number of L0 Signatures") + theme(axis.text.x = element_blank())

ggsave("../paper/plots/L0_by_fun.pdf",width=10,height=4)

plot_L1_df %>%
  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
  geom_line() + labs(title = "Number of Function Signatures per Function (L1)",
                                    x = "",
                                    y = "Number of L1 Signatures") + theme(axis.text.x = element_blank())

ggsave("../paper/plots/L1_by_fun.pdf",width=10,height=4)
```

@@

```{r}
# L2 Time
psp_L2 <- read_csv("../data/partial_L1_L2.csv.gz")
psp_L2 <- psp_L2[!psp_L2$pkg == "SUBTYPE",]
```

```{r}
# Plot it
pf_L2 <- psp_L2 %>% mutate(pkgfun = paste(pkg, fun, sep="::")) %>% select(pkgfun)

pf_L2 %>% mutate(num_sigs = 1) %>% group_by(pkgfun) %>% summarize(num_sigs=sum(num_sigs)) -> plot_L2

plot_L2 %>% arrange(num_sigs) %>% mutate(pkgfun_o = factor(pkgfun, pkgfun)) -> plot_L2_df

L2_perc_poly <- round(100 * nrow(plot_L2_df[plot_L2_df$num_sigs > 1, ] )/nrow(plot_L2_df), 1)

plot_L2_df %>%
  ggplot(aes(x=pkgfun_o, y=num_sigs, group=1)) +
  geom_line() + labs(title = "Number of Function Signatures per Function (L2)",
                                    x = "",
                                    y = "Number of L2 Signatures") + theme(axis.text.x = element_blank())

ggsave("../paper/plots/L2_by_fun.pdf",width=10,height=4)
```

```{r}
# L2 collapsed?
# what <- read_csv("../data/partial_L1_sanitized_class_collapsed.csv.gz")
what <- read_csv("../data/partial_L1_collapsed_L2_new.csv.gz")
```

### 2019-04-05 paper/main.tex
2/3 plots
last plot, modified discussion, still plugging away at it

@@

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
\begin{table}[ht]
\label{tab:L0top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 69806 & 26.3\% \\ 
  \l\ & 55311 & 20.8\% \\ 
  \C & 27789 & 10.5\% \\ 
  \L & 26230 & 9.9\% \\ 
  \sF & 16806 & 6.3\% \\ 
  \sN & 13976 & 5.3\% \\ 
  unevaled & 9125 & 3.4\% \\ 
  \I & 8971 & 3.4\% \\ 
  \sS & 4748 & 1.8\% \\ 
  \D, \I & 3256 & 1.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\end{table}

We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
Recall that in {\bf L0}, data frames will appear as lists.
As for ``unevaled'', first recall the discussion that all function arguments are wrapped in promises in R from \AT{where?}.
The ``unevaled'' signature appears whenever an argument was provided but unused, and one might wonder why such a signature might be so common.
As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, so it will appear as ``unevaled''.
The magrittr package's pipe \%$>$\% function is an example of this.
\AT{Move magrittr discussion up.}

One of our chief goals is to generate useful function signatures.
After tracing and capturing unique signatures from function call sites, \AT{recall} that we collapse some of the more obvious ones together.
After this, we are curious to know: {\it how many functions have more than one signature?}
Table~\ref{tab:L0funcounts} has the answer.

\begin{table}[ht]
\label{tab:L0funcounts}
\centering
\begin{tabular}{llr}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 46K & 83\% \\ 
  Total Seen & 56K & --- \\ 
   \hline
\end{tabular}
\caption{Function polymorphism in L0.}
\end{table}

Here, we see that the vast majority of functions are monomorphic in L0.
To choosing how to express the polymorphicity of a function, there are two main options one might consider:
First, having several separate whole-function signatures, e.g. \D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g. (\D $\cup$ \C) $\rightarrow$ \C.
Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument signatures.
To get an idea of how expressing polymorphism at the argument level would look, we ask: {\it what are the most frequently occurring polymorphic argument signatures?}
Table~\ref{tab:L0toppoly} has the answer.

@@

Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\label{tab:L1top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 48815 & 18.4\% \\ 
  \D & 27367 & 10.3\% \\ 
  \sL & 24284 & 9.2\% \\ 
  \sC & 22025 & 8.3\% \\ 
  \l & 20062 & 7.6\% \\ 
  \sF & 18480 & 7\% \\ 
  \M{D} & 14315 & 5.4\% \\ 
  \sN & 14016 & 5.3\% \\ 
  unevaled & 9093 & 3.4\% \\ 
  \C & 7731 & 2.9\% \\
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\end{table}

@@

Now that we have a picture of how arguments look in L1, let's see: {\it how many L1 functions have more than one type signature?}
The answer can be found in Table~\ref{tab:funcountsL1}.

\begin{table}[ht]
\label{tab:funcountsL1}
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 44K & 79.6\% \\ 
  Total Seen & 56K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in L1.}
\end{table}

@@

As you can see, the most frequent signatures in {\bf L1} are represented here.
The inclusion of class and attribute information has changed the counts slightly.
One tidbit of interest here is that monomorphic named lists are more common than their unnamed counterparts.
The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.
%On that note, we wonder: {\it what are most frequent argument signatures with an interesting class?}
%Table~\ref{tab:L2top10argwclass} has the answer.

%% latex table generated in R 3.5.1 by xtable 1.8-3 package
%% Fri Apr  5 02:38:12 2019
%\begin{table}[ht]
%\label{tab:L2top10argwclass}
%\centering
%\begin{tabular}{lllrl}
%  \hline
%Type & Class & Attr Ptrn & Count & \% of All Signatures \\ 
%  \hline
%  \sE & environment & \{\} & 704 & 0.3\% \\ 
%  language & formula & \{.Environment\} & 544 & 0.2\% \\ 
%  \D & array & \{dim\} & 395 & 0.1\% \\ 
%  externalptr & externalptr & \{\} & 301 & 0.1\% \\ 
%  \sF & standardGeneric & \{default,...,srcref,valueClass\} & 255 & 0.1\% \\ 
%  \df & tbl\_df,tbl & \{names,row.names\} & 239 & 0.1\% \\ 
%  \sF & standardGeneric & \{default,...,valueClass\} & 239 & 0.1\% \\ 
%  \l & phylo & \{names,order\} & 199 & 0.1\% \\ 
%  \l & lm & \{names\} & 175 & 0.1\% \\ 
%  \l & lvm & \{names\} & 157 & 0.1\% \\ 
%   \hline
%\end{tabular}
%\caption{Most frequent argument signatures with class in {\bf L2}.}
%\end{table}
%
%Table~\ref{tab:L2top10argwclass} does indeed showcase some interesting classes.
%The three builtin classes which we have not yet seen are environment, formula, and array.
%An environments is a list which has a references to a parent environment (another list with environment class).
%A formula is a \AT{small explanation}.
%An array is a generalization of a matrix in R, where an array can have more than 2 dimensions.
%Each of the other classes pictured here are defined in some package, and are not part of the base language.
%An interesting fact about this table is that all of the pictures signatures are monomorphic.

Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
For this account of polymorphism, we deem an argument to be polymorphic if it was polymorphic in type, class, or attribute pattern.
Consider Table~\ref{tab:funcountsL2}.
\AT{This is where the per fun plot should go, will do.}

\begin{table}[ht]
\label{tab:funcountsL2}
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 44K & 79.6\% \\ 
  Total Seen & 56K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in {\bf L2}.}
\end{table}

\AT{Connect, talk about increase/decrease in poly.}

@@

\AT{Will probably change.}

>>>>

\newcommand{\LZEROPERCPOLY}{17\%\xspace}
\newcommand{\LONEPERCPOLY}{20.4\%\xspace}
\newcommand{\LTWOPERCPOLY}{23\%\xspace}

@@

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
\begin{table}[ht]
\label{tab:L0top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 69806 & 26.3\% \\ 
  \l\ & 55311 & 20.8\% \\ 
  \C & 27789 & 10.5\% \\ 
  \L & 26230 & 9.9\% \\ 
  \sF & 16806 & 6.3\% \\ 
  \sN & 13976 & 5.3\% \\ 
  \ANY & 9125 & 3.4\% \\ 
  \I & 8971 & 3.4\% \\ 
  \sS & 4748 & 1.8\% \\ 
  \D, \I & 3256 & 1.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\end{table}

We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
Recall that in {\bf L0}, data frames will appear as lists.
The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, and if this metaprogramming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming.

One of our chief goals is to generate useful function signatures.
After tracing and capturing unique signatures from function call sites, \AT{recall} that we collapse some of the more obvious ones together.
After this, we are curious to know: {\it how many functions have more than one signature?}
Figure~\ref{fig:L0funcounts} has the answer with a breakdown by function.

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L0_by_fun}
\caption{Record of Function Polymorphism in {\bf L0}. \LZEROPERCPOLY of functions are polymorphic.}
\label{fig:L0funcounts}\end{center}
\end{figure}

Here, we see that the vast majority of functions are monomorphic in L0.
Indeed, there are a few functions with a disproportionately large number of signatures.
One such function is the magrittr package's pipe \%$>$\%:

In choosing how to express the polymorphicity of a function, there are two main options one might consider:
First, having several separate whole-function signatures, e.g. \D $\rightarrow$ \D and \C $\rightarrow$ \D, or alternatively expressing the polymorphism at the argument level with a union type, e.g. (\D $\cup$ \C) $\rightarrow$ \C.
Earlier in Table~\ref{tab:L0top10arg} we saw the most frequently occurring argument signatures.
To get an idea of how expressing polymorphism at the argument level would look, we ask: {\it what are the most frequently occurring polymorphic argument signatures?}
Table~\ref{tab:L0toppoly} has the answer.

@@

Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\label{tab:L1top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 48815 & 18.4\% \\ 
  \D & 27367 & 10.3\% \\ 
  \sL & 24284 & 9.2\% \\ 
  \sC & 22025 & 8.3\% \\ 
  \l & 20062 & 7.6\% \\ 
  \sF & 18480 & 7\% \\ 
  \M{D} & 14315 & 5.4\% \\ 
  \sN & 14016 & 5.3\% \\ 
  \ANY & 9093 & 3.4\% \\ 
  \C & 7731 & 2.9\% \\
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\end{table}

@@

Now that we have a picture of how arguments look in L1, let's see: {\it how many L1 functions have more than one type signature?}
The answer can be found in Figure~\ref{fig:L1funcounts}.

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L1_by_fun}
\caption{Record of Function Polymorphism in {\bf L1}. \LONEPERCPOLY of functions are polymorphic.}
\label{fig:L1funcounts}\end{center}
\end{figure}

@@

As you can see, the most frequent signatures in {\bf L1} are represented here.
The inclusion of class and attribute information has changed the counts slightly.
One tidbit of interest here is that monomorphic named lists are more common than their unnamed counterparts.
The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.

Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
For this account of polymorphism, we deem an argument to be polymorphic if it was polymorphic in type, class, or attribute pattern.
Consider Figure~\ref{fig:L2funcounts}.

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{L2_by_fun}
\caption{Record of Function Polymorphism in {\bf L2}. \LTWOPERCPOLY of functions are polymorphic.}
\label{fig:L2funcounts}\end{center}
\end{figure}

@@

\AT{Make into subsection at the end of analysis? Or in global results, as a takeaway.}
